---
title: "ABC-SMC"
format: html
editor: visual
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(zellkonverter)
  library(Matrix)
  library(SingleCellExperiment)
  library(ggplot2)
  library(gridExtra)
})

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
```

We want to emulate the scRNA-seq data-generating process:

$$
z_{ij}\ \xrightarrow[\text{scRNA-seq measurement}]{t_{ij}}\ x_{ij}.
$$

Let $i=1,\dots,N$ index cells and $j=1,\dots,J$ index genes. There are two steps:

1.  Generate latent true expression $z_{ij}\ge 0$ from a biological model (with on/off state and biological variability).

2.  Generate observed scRNA-seq expression (UMI counts) $x_{ij}\ge0$ from an scRNA-seq measurement model (with technical variability).

## **1. Simulator Model**

### **1.1 Overview**

Hyperparameters: $$
\lambda=(\theta,\gamma,\rho,\alpha,\sigma^2,\alpha_\psi,\beta_\psi,\alpha_\phi,\beta_\phi).
$$

Let the total number of genes be $J$, where genes $1,\dots,K$ are transcription factors (TFs) that act as regulators. All $J$ genes may serve as regulatory targets, while genes $K+1,\dots,J$ are non-TF genes.

Let $u_{ij} \in [0, 1]$ act as the continuous perturbation indicator for gene $j$ in cell $i$, representing the degree of knockdown. Within each simulation, compute the perturbation-aware regulatory means:

$$
m_{ij}=\exp\left(\sum_{k=1}^K \beta_{ikj}^{\prime}a_{ik}\right) 
$$ with TF activities of regulator $k$:\
$$
a_{ik}\sim\text{Gamma}(\alpha,\alpha)
$$ and regulatory effect $$
\beta_{ikj}^{\prime} = \beta_{kj} \ (1 - u_{ik}), \quad \beta_{kj}= \delta_{kj} \dot{\beta}_{kj}, \quad \dot{\beta}_{kj}\sim\mathcal N(0,\sigma^2).
$$

$$
\delta_{kj}\mid\psi_k,\phi_j \sim\operatorname{Bernoulli}(\psi_k\phi_j), \quad \psi_k \sim\operatorname{Beta}(\alpha_\psi,\beta_\psi), \quad \phi_j \sim\operatorname{Beta}(\alpha_\phi,\beta_\phi)
$$

Define gene-specific gene-on probabilities: $$
\pi_j\sim \text{Beta}(\theta,\ (1-\theta)),\quad \theta \in (0,1)
$$ Then, generating the latent state and applying the target perturbation mask: $$
v_{ij} \ | \ \pi_j\sim \text{Bernoulli}(\pi_j),\qquad z_{ij}= \begin{cases} 0,& v_{ij}=0\\ \text{Gamma}(m_{ij}\gamma,\gamma),& v_{ij}=1 \end{cases}, \qquad z_{ij}^{\prime} = z_{ij} \ (1 - u_{ij})
$$ $$
t_{ij}\sim\text{Gamma}(\rho,\rho),\qquad x_{ij}\mid z_{ij}^{\prime},t_{ij}\sim\text{Poisson}(t_{ij}z_{ij}^{\prime})
$$

### **1.2 Biological Model for Latent True Expression** $z_{ij}$

#### **1.2.1 Regulatory model for** $m_{ij}$

To account for continuous perturbations, we define a cell-specific masked regulatory effect $\beta_{ikj}^{\prime}$: $$
\beta_{ikj}^{\prime} = \beta_{kj} \  (1 - u_{ik}), \quad i=1, \dots, N.
$$

This ensures that if regulator $k$ is partially or fully knocked down in cell $i$, its regulatory effect on downstream genes scales proportionally. We then define the regulatory mean as:

$$
m_{ij}=\exp\left(\sum_{k=1}^K \beta_{ikj}^{\prime}a_{ik}\right)  
$$

with:

-   $a_{ik}\ge 0$ denoting the TF activities of regulator $k$ in cell $i$, such that: $$
      a_{ik}\sim \text{Gamma}(\alpha,\alpha)\quad\Rightarrow\quad \mathbb E[a_{ik}]=1,\ \quad \text{Var}(a_{ik})=\frac{1}{\alpha}.
    $$

-   $\delta_{kj}$ denoting the binary adjacency matrix, such that:

    $$
    \delta_{kj}\mid\psi_k,\phi_j \sim\operatorname{Bernoulli}(\psi_k\phi_j),
    $$

    where the regulator-specific and target-specific edge propensities satisfy $$\psi_k\sim\operatorname{Beta}(\alpha_\psi,\beta_\psi),
    \qquad k=1,\ldots,K, \quad \text{and} \quad \phi_j\sim\operatorname{Beta}(\alpha_\phi,\beta_\phi),
    \qquad j=1,\ldots,J.$$

    Here,

    $$
    \delta_{kj} = \begin{cases} 1, & \text{regulator } k \text{ regulates target } j, \\ 0, & \text{otherwise.} \end{cases}
    $$

-   $\dot{\beta}_{kj}$ denoting the signed regulatory effect of regulator $k$ on gene $j$, such that: $$
    \dot{\beta}_{kj}\sim\mathcal N(0,\sigma^2).
    $$

    This yields the regulatory effect: $$
    \beta_{kj}= \delta_{kj}\dot{\beta}_{kj}
    $$

#### **1.2.2 On/Off States** $v_{ij}$ **and Perturbation Scaling**

We conceptualize the latent true expression $z_{ij}^{\prime}$ as being governed by the intersection of a binary biological switch and a continuous experimental knockdown. This ensures that a gene only displays expression if it is biologically on, and that the magnitude of this expression is modulated by the efficiency of the CRISPR perturbation.

##### **1.2.2.1 The Biological Switch (**$v_{ij}$**)**

This represents the natural transcriptional state of gene $j$ in cell $i$. We introduce a global hyperparameter $\theta \in (0,1)$ to define the gene-specific on-probability $\pi_j$:

$$
\pi_j\sim \text{Beta}(\theta,\ (1-\theta))\quad\Rightarrow\quad \mathbb E[\pi_j]=\theta,\ \quad 
\mathrm{Var}(\pi_j)=\frac{\theta(1-\theta)}{2}.
$$

The biological state is then sampled as:

$$
v_{ij}\sim \mathrm{Bernoulli}(\pi_j).
$$

-   $v_{ij} = 1$: The gene is biologically on.

-   $v_{ij} = 0$: The gene is biologically off.

##### **1.2.2.2 The Perturbation Scaling (**$u_{ij}$**)**

Unlike a binary knockout, the experimental intervention is modeled as a continuous knockdown fraction $u_{ij} \in [0, 1]$. This represents the proportion of expression lost due to the perturbation:

-   $u_{ij} = 0$: No perturbation (100% of biological activity remains).

-   $0 < u_{ij} < 1$: Partial knockdown (e.g., $u_{ij} = 0.9$ implies 10% residual activity).

-   $u_{ij} = 1$: Complete knockout (0% of biological activity remains).

##### **1.2.2.3 Latent Expression Logic**

To implement the perturbation on the target gene, we define the masked latent expression $z_{ij}^{\prime}$ by scaling the theoretical biological latent state $z_{ij}$ by the residual activity $(1 - u_{ij})$:

1.  Case 1: $z_{ij}^{\prime} = 0$ if $v_{ij} = 0$

    If the biological switch is off, the gene cannot be expressed regardless of the perturbation state.

$$
\Pr(z_{ij}^{\prime}=0\mid v_{ij}=0)=1,\qquad \Pr(z_{ij}^{\prime}\neq 0\mid v_{ij}=0)=0
$$

2.  Case 2: $z_{ij}^{\prime} \ge 0$ if $v_{ij} = 1$

    If the gene is biologically on, its latent expression is sampled from a Gamma distribution centered around the regulatory mean $m_{ij}$, then scaled by the knockdown factor $(1 - u_{ij})$:

    $$
    z_{ij}^{\prime}\mid (v_{ij}=1)\ \sim\ \text{Gamma}(m_{ij}\gamma,\ \gamma) \ (1 - u_{ij})
    $$

    where $\gamma > 0$ represents biological variability. Under this parameterization, the conditional mean and variance of the masked expression are:

$$
\mathbb{E}[z_{ij}^{\prime}\mid v_{ij}=1]=m_{ij}(1 - u_{ij}),\qquad \text{Var}(z_{ij}^{\prime}\mid v_{ij}=1)=\frac{m_{ij}(1 - u_{ij})^2}{\gamma}.
$$

### **1.3 Observed scRNA-seq Count** $x_{ij}$

We map the masked latent true expression $z_{ij}^{\prime}$ to the observed UMI count: $$
x_{ij}\mid z_{ij}^{\prime},t_{ij}\sim\text{Poisson}(t_{ij}z_{ij}^{\prime})
$$ with $t_{ij}$ denoting technical variability, which is modeled by a Gamma prior: $$
  t_{ij}\sim\text{Gamma}(\rho,\rho)\quad\Rightarrow\quad 
  \mathbb{E}[t_{ij}] = 1,\qquad \mathrm{Var}(t_{ij}) = \frac{1}{\rho}. 
$$

Here, $\rho$ controls the magnitude of the technical noise. Because the perturbation is directly baked into $z_{ij}^{\prime}$, any cell where a gene is knocked out will have $z_{ij}^{\prime}=0$. Consequently, the Poisson rate ($t_{ij} \cdot 0$) becomes exactly 0, guaranteeing that the observed UMI count $x_{ij} = 0$ for that knockout target.

## **2. Summary Statistics**

### **2.1 Entry-specific summaries**

Given an observed UMI count matrix $X_{\text{obs}} \in \mathbb{R}^{N\times J}$, we summarize the marginal distribution of entrywise counts by considering the vector of all entries

$$
x_\text{all} = (X_{ij} : i=1,...,N,\ j=1,...,J).
$$

Because scRNA-seq matrices are often highly zero-inflated, we additionally summarize the distribution of positive counts only:

$$
x_\text{nonzero} = (X_{ij} : X_{ij} > 0).
$$

### **2.2 Cell-specific summaries**

To preserve variation across cells on a scale comparable across datasets, we use the mean UMI count per gene in each cell:

$$
\bar C_i = \frac{1}{J}\sum_{j=1}^J X_{ij},\qquad i=1,\dots,N.
$$

We also summarize the fraction of detected genes (nonzero genes) per cell:

$$
\bar D_i = \frac{1}{J}\sum_{j=1}^{J} \mathbf{1}\{X_{ij} > 0\}, \qquad i=1,\dots,N,
$$

and let $C=(\bar C_i)_{i=1}^N$ and $D=(\bar D_i)_{i=1}^N$.

### **2.3 Gene-specific summaries**

To also preserve variation across genes on a comparable scale, we use the mean UMI count per cell for each gene:

$$
\bar G_j = \frac{1}{N}\sum_{i=1}^N X_{ij},\qquad j=1,\dots,J.
$$

We also summarize the fraction of cells expressing each gene:

$$
\bar M_j = \frac{1}{N}\sum_{i=1}^N \mathbf{1}\{X_{ij}>0\},\qquad j=1,\dots,J,
$$

and let $G=(\bar G_j)_{j=1}^J$ and $M=(\bar M_j)_{j=1}^J$.

### **2.4 Quantile summaries**

We combine these distributions into a single summary vector $S(X)$ by taking selected quantiles from each component. Let

$$
P= \{0.1, 0.2, 0.3,0.4,0.5, 0.6, 0.7,0.8,0.9\}.
$$

For any vector $v$, let $Q_v(p)$ denote the $p$-th quantile of the entries of $v$, and let $Q_v(P)=(Q_v(p):p\in P)$. The final summary vector is

$$
S(X) = \Bigl( Q_{x_{\text{all}}}(P),\; Q_{x_{\text{nonzero}}}(P),\; Q_{C}(P),\; Q_{D}(P),\; Q_{G}(P),\; Q_{M}(P) \Bigr).
$$

For a simulated dataset $X_\text{sim}$, we compute $S_\text{sim} = S(X_\text{sim})$ and compare it to the observed summary $S_\text{obs}=S(X_\text{obs})$ within the Approximate Bayesian Computation procedure.

## **3. Prior**

We start by choosing a prior over hyperparameters

$$
\lambda=(\theta,\gamma,\rho,\alpha,\sigma^2,\alpha_\psi,\beta_\psi,\alpha_\phi,\beta_\phi).
$$

to reflect plausible ranges for each mechanism in the simulator (on/off, biological variability, technical variability, activity variability, effect sizes, regulator-specific edge propensity, and target-specific edge propensity).

-   Overall fraction of probability that a gene is on

$$
\theta \sim \text{Uniform}(0.05,\;0.95).
$$

-   Biological Gamma concentration

$$
\log \gamma \sim \text{Uniform}(\log 1.5,\;\log 25).
$$

-   Technical Gamma concentration

$$
\log \rho \sim \text{Uniform}(\log 10,\;\log 60).
$$

-   Multiplicative TF-activity Gamma concentration

$$
\log \alpha \sim \text{Uniform}(\log 2,\;\log 20).
$$

-   Effect variance

$$
\log \sigma^2 \sim \text{Uniform}(0.05,\;0.15)
$$

-   Regulator-specific edge-propensity Beta shape hyperparameters

    $$
    \log\alpha_\psi\sim\operatorname{Uniform}(\log 0.1,\log 5),
    $$

    $$
    \log\beta_\psi\sim\operatorname{Uniform}(\log 0.1,\log 10).
    $$

-   Target-specific edge-propensity Beta shape hyperparameters

    $$
    \log\alpha_\phi\sim\operatorname{Uniform}(\log 0.1,\log 5),
    $$

    $$
    \log\beta_\phi\sim\operatorname{Uniform}(\log 0.1,\log 10).
    $$

For each sampled $\lambda$, we generate a simulated dataset ${X}_{\text{sim}}$ using Sections 1-2.

## **4. ABC-SMC**

We infer $\lambda=(\theta,\gamma,\rho,\alpha,\sigma^2,\alpha_\psi,\beta_\psi,\alpha_\phi,\beta_\phi)$ using an Approximate Bayesian Computation Sequential Monte Carlo (ABC-SMC) procedure, following Castro (2025). The goal is to approximate the posterior distribution of $\lambda$ by evolving a sequence of particle populations whose simulated summary statistics become progressively closer to the observed summaries.

At each stage, we compare the simulated and observed summaries using the Euclidean discrepancy

$$
d(\lambda)\;=\;\bigl\|S_{\text{sim}}-S_{\text{obs}}\bigr\|_2,
$$

where $S_{\text{sim}}=S(X_{\text{sim}})$ and $S_{\text{obs}}=S(X_{\text{obs}})$.

We use

$$W=200$$

particles per population and run $T=5$ ABC-SMC populations.

### **4.1 First population**

For the first population, we generate particles using rejection ABC. First, we draw a pilot sample of hyperparameter values from the prior and use their distances to determine the initial tolerance $\epsilon_1$. Specifically, we simulate a pilot set of size

$$
Q_{\text{pilot}} = 2,000,
$$

compute the corresponding distances, and set $\epsilon_1$ to the 0.25-quantile of the pilot distance distribution, that is,

$$
\varepsilon_1 = Q_{0.25}(d^{(1)},\dots,d^{(Q_{\text{pilot}})}).
$$

We then repeatedly:

1.  sample $\lambda \sim \pi(\lambda)$ from the prior,

2.  simulate $X_{\text{sim}}$,

3.  compute $S_{\text{sim}}$ and $d(\lambda)$,

4.  accept $\lambda$ if

    $$
    d(\lambda)\le \varepsilon_1.
    $$

This process continues until exactly $W=200$ particles are accepted. All particles in the first population are assigned equal weights.

### **4.2 Later populations**

For each later population $t=2,\dots,T$, we update the tolerance automatically using the accepted distances $d_{t-1}^{(1)},\dots,d_{t-1}^{(W)}$ from the previous population $t-1$. Specifically, we set

$$
\varepsilon_t = Q_{0.25}(d_{t-1}^{(1)},\dots,d_{t-1}^{(W)}),
$$

so that the tolerance is decreased adaptively across populations rather than fixed in advance.

To generate the $t$-th population, we repeatedly:

1.  sample one particle from the previous population with probability proportional to its weight,

2.  perturb it using a multivariate Gaussian kernel,

3.  reject immediately if the perturbed hyperparameter falls outside the allowed prior ranges,

4.  simulate $X_{\text{sim}}$ from the proposed hyperparameter,

5.  compute $S_{\text{sim}}$ and $d(\lambda)$,

6.  accept the proposal if

    $$
    d(\lambda)\le \varepsilon_t.
    $$

We continue until exactly $W=200$ particles are accepted in that population.

### **4.3 Gaussian perturbation kernel and weights**

For population $t \ge 2$, we first select a parent particle $\lambda^*$ from the previous population with probability proportional to its weight, and then perturb it using a multivariate Gaussian random-walk kernel on the transformed hyperparameter scale. Let $\Sigma_{t-1}$ denote the weighted empirical covariance matrix of the transformed particles in population $t-1$. We propose a new particle $\lambda$ from

$$
K_t(\lambda \mid \lambda^*) = \mathcal N(\lambda^*,\, 2\Sigma_{t-1}),
$$

so that the proposal is centered at the parent particle $\lambda^*$, and its covariance is tuned to the spread of the previous particle population. The factor 2 slightly enlarges the proposal covariance relative to the empirical particle covariance, allowing broader local exploration.

After acceptance, particles are assigned importance weights. In the first population, all accepted particles are assigned equal weights. For populations $t \ge 2,$ the weight of accepted particle $\lambda_t^{(m)}$ is updated by

$$
w_t^{(m)} \propto \frac{\pi(\lambda_t^{(m)})}{ \sum_{\ell=1}^W w_{t-1}^{(\ell)}\,K_t\!\left(\lambda_t^{(m)} \mid \lambda_{t-1}^{(\ell)}\right) }.$$

where $\pi(\lambda_t^{(m)})$ is the prior density of the accepted particle, $w_{t-1}^{(\ell)}$ is the normalized weight of particle $\ell$ in the previous population, and $K_t(\lambda_t^{(m)} \mid \lambda_{t-1}^{(\ell)})$ is the proposal density from previous particle $\ell$ to the accepted particle $\lambda_t^{(m)}$. The weights are then normalized to sum to one within each population.

### **4.4 ESS and resampling**

At each population, we monitor the effective sample size (ESS),

$$
\mathrm{ESS} = \frac{1}{\sum_{m=1}^W (w_t^{(m)})^2},
$$

as a diagnostic for particle degeneracy. If the ESS falls below half of the particle count, we resample the accepted particles and reset their weights to be uniform.

### **4.5 Final posterior and posterior predictive checks**

After the final population $T=5,$ we take the last weighted particle population as the ABC-SMC approximation to the posterior distribution of

$$
\lambda=(\theta,\gamma,\rho,\alpha,\sigma^2,\alpha_\psi,\beta_\psi,\alpha_\phi,\beta_\phi).
$$

Using the accepted particles from the final population, we generate posterior predictive datasets and compare their distributions with $X_{\text{obs}}$.

```{r}
norman <- readH5AD("/Users/zhangrui/Desktop/precmat/Norman_2019.h5ad", 
                   reader = "R", use_hdf5 = TRUE)

# Single-gene perturbations = regulators
pert_all <- as.character(colData(norman)$perturbation_name)
regulator_gene_names <- sort(unique(pert_all[!is.na(pert_all) & pert_all != "control" &
                                               !grepl("\\+", pert_all)]))

# Control-cell raw UMI matrix
sce_control <- norman[, colData(norman)$perturbation_name == "control"]
X_real <- assay(sce_control, "counts")

# Match regulators to expression matrix
regulator_gene_idx <- match(regulator_gene_names, rownames(X_real))
regulator_gene_idx <- regulator_gene_idx[!is.na(regulator_gene_idx)]

# Subsample
set.seed(1)

N_use <- 300
J_use <- 200
K_use <- 40

regulator_use <- sample(regulator_gene_idx, K_use)

non_regulator <- setdiff(seq_len(nrow(X_real)), regulator_gene_idx)
other_genes <- sample(non_regulator, J_use - K_use)

genes_use <- sample(c(regulator_use, other_genes))
cells_use <- sample(seq_len(ncol(X_real)), N_use)

# N x J raw UMI matrix
X_obs <- as.matrix(t(X_real[genes_use, cells_use, drop = FALSE]))

cat("Observed matrix dimension N x J:", dim(X_obs), "\n")
```

```{r}
# 1) Summary statistics for ABC
summaries <- function(X) {

  x_all <- as.vector(X)
  x_nonzero <- x_all[x_all > 0 & !is.na(x_all)]

  if (length(x_nonzero) == 0 || any(!is.finite(x_all))) {
    return(rep(NA, 54))
  }

  C_i <- rowMeans(X)
  D_i <- rowMeans(X > 0)
  G_j <- colMeans(X)
  M_j <- colMeans(X > 0)

  probs_use <- c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9)

  c(
    quantile(x_all, probs = probs_use, na.rm = TRUE, names = FALSE),
    quantile(x_nonzero, probs = probs_use, na.rm = TRUE, names = FALSE),
    quantile(C_i, probs = probs_use, na.rm = TRUE, names = FALSE),
    quantile(D_i, probs = probs_use, na.rm = TRUE, names = FALSE),
    quantile(G_j, probs = probs_use, na.rm = TRUE, names = FALSE),
    quantile(M_j, probs = probs_use, na.rm = TRUE, names = FALSE)
  )
}

S_obs <- summaries(X_obs)

# 2) ABC distance
distance_abc <- function(S_sim, S_obs) {
  d <- sqrt(sum((S_sim - S_obs)^2))

  # If distance is NA, return Infinity to force rejection
  if (!is.finite(d)) {
    return(Inf)
  }

  d
}

# 3) Prior ranges
prior_range <- list(
  theta = c(0.05, 0.95),
  gamma = c(1.5, 25),
  rho = c(10, 60),
  alpha = c(2, 20),
  sigma2 = c(0.05, 0.15),
  alpha_phi = c(0.1, 5),
  beta_phi = c(0.1, 10),
  alpha_psi = c(0.1, 5),
  beta_psi = c(0.1, 10)
)

# Sample one hyperparameter from its prior
sample_lambda <- function() {
  list(
    # Uniform prior
    theta = runif(1, min = prior_range$theta[1], max = prior_range$theta[2]),

    # Log-uniform priors
    gamma = exp(runif(1, min = log(prior_range$gamma[1]), max = log(prior_range$gamma[2]))),
    rho = exp(runif(1, min = log(prior_range$rho[1]), max = log(prior_range$rho[2]))),
    alpha = exp(runif(1, min = log(prior_range$alpha[1]), max = log(prior_range$alpha[2]))),
    sigma2 = exp(runif(1, min = log(prior_range$sigma2[1]), max = log(prior_range$sigma2[2]))),
    alpha_phi = exp(runif(1, min = log(prior_range$alpha_phi[1]), 
                          max = log(prior_range$alpha_phi[2]))),
    beta_phi = exp(runif(1, min = log(prior_range$beta_phi[1]), 
                         max = log(prior_range$beta_phi[2]))),
    alpha_psi = exp(runif(1, min = log(prior_range$alpha_psi[1]), 
                          max = log(prior_range$alpha_psi[2]))),
    beta_psi = exp(runif(1, min = log(prior_range$beta_psi[1]),
                         max = log(prior_range$beta_psi[2])))
  )
}

# 4) Prior density
prior_density <- function(lambda) {

  inside_support <- (
    lambda$theta >= prior_range$theta[1] &&
      lambda$theta <= prior_range$theta[2] &&
    lambda$gamma >= prior_range$gamma[1] &&
      lambda$gamma <= prior_range$gamma[2] &&
    lambda$rho >= prior_range$rho[1] &&
      lambda$rho <= prior_range$rho[2] &&
    lambda$alpha >= prior_range$alpha[1] &&
      lambda$alpha <= prior_range$alpha[2] &&
    lambda$sigma2 >= prior_range$sigma2[1] &&
      lambda$sigma2 <= prior_range$sigma2[2] &&
    lambda$alpha_phi >= prior_range$alpha_phi[1] &&
      lambda$alpha_phi <= prior_range$alpha_phi[2] &&
    lambda$beta_phi >= prior_range$beta_phi[1] &&
      lambda$beta_phi <= prior_range$beta_phi[2] &&
    lambda$alpha_psi >= prior_range$alpha_psi[1] &&
      lambda$alpha_psi <= prior_range$alpha_psi[2] &&
    lambda$beta_psi >= prior_range$beta_psi[1] &&
      lambda$beta_psi <= prior_range$beta_psi[2]
  )

  if (!inside_support) {
    return(0)
  }

  d_theta <- 1 / (prior_range$theta[2] - prior_range$theta[1])
  d_log_gamma <- 1 / (log(prior_range$gamma[2]) - log(prior_range$gamma[1]))
  d_log_rho <- 1 / (log(prior_range$rho[2]) - log(prior_range$rho[1]))
  d_log_alpha <- 1 / (log(prior_range$alpha[2]) - log(prior_range$alpha[1]))
  d_log_sigma2 <- 1 / (log(prior_range$sigma2[2]) - log(prior_range$sigma2[1]))
  d_log_alpha_phi <- 1 / (log(prior_range$alpha_phi[2]) - log(prior_range$alpha_phi[1]))
  d_log_beta_phi <- 1 / (log(prior_range$beta_phi[2]) - log(prior_range$beta_phi[1]))
  d_log_alpha_psi <- 1 / (log(prior_range$alpha_psi[2]) - log(prior_range$alpha_psi[1]))
  d_log_beta_psi <- 1 / (log(prior_range$beta_psi[2]) - log(prior_range$beta_psi[1]))

  d_theta * d_log_gamma * d_log_rho * d_log_alpha * d_log_sigma2 * d_log_alpha_phi * d_log_beta_phi * d_log_alpha_psi * d_log_beta_psi
}

# check hyperparameter vector inside support of prior
in_prior_support <- function(theta) {
  prior_density(theta) > 0
}

# 5) Simulator
simulate_X_from_lambda <- function(lambda, N, J, K) {

  # hyperparameters
  theta <- lambda$theta
  gamma <- lambda$gamma
  rho <- lambda$rho
  alpha <- lambda$alpha
  sigma2 <- lambda$sigma2
  alpha_phi <- lambda$alpha_phi
  beta_phi <- lambda$beta_phi
  alpha_psi <- lambda$alpha_psi
  beta_psi <- lambda$beta_psi

  # (a) TF activity a_ik
  a_ik <- matrix(rgamma(N * K, shape = alpha, rate = alpha), nrow = N, ncol = K)

  # (b) TF -> gene effects: beta_kj ~ Normal(0, sigma2)
  beta_kj <- matrix(rnorm(K * J, mean = 0, sd = sqrt(sigma2)), nrow = K, ncol = J)

  # (c) delta_kj ~ Bernoulli(phi_j * psi_k)
  phi_j <- rbeta(J, shape1 = alpha_phi, shape2 = beta_phi)
  psi_k <- rbeta(K, shape1 = alpha_psi, shape2 = beta_psi)
  prob_kj <- outer(psi_k, phi_j, "*")
  delta_kj <- matrix(rbinom(K * J, size = 1, prob = as.vector(prob_kj)), nrow = K, ncol = J)
  
  beta_kj <- beta_kj * delta_kj

  # (d) Mean expression: m_ij = exp(sum_k a_ik * beta_kj)
  m_ij <- exp(a_ik %*% beta_kj)

  # (e) Gene-specific on-probabilities
  shape1 <- theta
  shape2 <- (1 - theta)
  pi_j <- rbeta(J, shape1 = shape1, shape2 = shape2)

  # On/off states: v_ij ~ Bernoulli(pi_j)
  v_ij <- matrix(
    rbinom(N * J, size = 1, prob = rep(pi_j, each = N)),
    nrow = N, ncol = J
  )

  # (f) Biological expression
  z_ij <- matrix(0, nrow = N, ncol = J)
  idx_on <- which(v_ij == 1, arr.ind = TRUE)

  if (nrow(idx_on) > 0) {
    shape_vec <- m_ij[idx_on] * gamma
    z_ij[idx_on] <- rgamma(nrow(idx_on), shape = shape_vec, rate = gamma)
  }

  # (g) Technical multiplicative noise: t_ij ~ Gamma(rho, rho)
  t_ij <- matrix(rgamma(N * J, shape = rho, rate = rho), nrow = N, ncol = J)

  # (h) Observation: X_ij | z_ij, t_ij ~ Poisson(t_ij * z_ij)
  lambda_ij <- t_ij * z_ij
  X_ij <- matrix(rpois(N * J, lambda = as.vector(lambda_ij)), nrow = N, ncol = J)

  list(
    X = X_ij,
    a = a_ik,
    beta = beta_kj,
    delta = delta_kj,
    m = m_ij,
    v = v_ij,
    z = z_ij,
    t = t_ij,
    pi_j = pi_j,
    phi_j = phi_j,
    psi_k = psi_k
  )
}
```

```{r}
# Convert lambda to numeric vector
lambda_to_vec <- function(lambda) {
  c(
    theta = lambda$theta,
    log_gamma = log(lambda$gamma),
    log_rho = log(lambda$rho),
    log_alpha = log(lambda$alpha),
    log_sigma2 = log(lambda$sigma2),
    log_alpha_phi = log(lambda$alpha_phi),
    log_beta_phi = log(lambda$beta_phi),
    log_alpha_psi = log(lambda$alpha_psi),
    log_beta_psi = log(lambda$beta_psi)
  )
}

# Convert back
vec_to_lambda <- function(v) {
  list(
    theta = unname(v["theta"]),
    gamma = exp(unname(v["log_gamma"])),
    rho = exp(unname(v["log_rho"])),
    alpha = exp(unname(v["log_alpha"])),
    sigma2 = exp(unname(v["log_sigma2"])),
    alpha_phi = exp(unname(v["log_alpha_phi"])),
    beta_phi = exp(unname(v["log_beta_phi"])),
    alpha_psi = exp(unname(v["log_alpha_psi"])),
    beta_psi = exp(unname(v["log_beta_psi"]))
  )
}

# Compute the weighted covariance matrix of the particles
weighted_covariance <- function(X, w) {
  mu <- colSums(X * w)
  Xc <- sweep(X, 2, mu, "-")
  crossprod(Xc * sqrt(w), Xc * sqrt(w))
}

# Stabilize covariance matrix so chol() does not fail
make_positive_definite <- function(Sigma, min_eig = 1e-8) {

  Sigma <- as.matrix(Sigma)
  Sigma[!is.finite(Sigma)] <- 0

  # Force symmetry
  Sigma <- (Sigma + t(Sigma)) / 2

  eig_vals <- eigen(Sigma, symmetric = TRUE, only.values = TRUE)$values
  min_val <- min(eig_vals, na.rm = TRUE)

  if (!is.finite(min_val)) {
    return(diag(min_eig, nrow(Sigma)))
  }

  if (min_val < min_eig) {
    diag(Sigma) <- diag(Sigma) + (min_eig - min_val)
  }

  # Force symmetry again after diagonal adjustment
  Sigma <- (Sigma + t(Sigma)) / 2

  Sigma
}

# Compute the log density of a multivariate Gaussian distribution
mvnorm_logpdf <- function(x, mean, Sigma) {

  Sigma <- make_positive_definite(Sigma, min_eig = 1e-8)

  p <- length(x)

  cholS <- tryCatch(
    chol(Sigma),
    error = function(e) {
      chol(make_positive_definite(Sigma, min_eig = 1e-6))
    }
  )

  z <- backsolve(cholS, x - mean, transpose = TRUE)

  out <- -0.5 * (
    p * log(2 * pi) +
      2 * sum(log(diag(cholS))) +
      sum(z^2)
  )

  if (!is.finite(out)) {
    return(-Inf)
  }

  out
}

# Take one old particle and perturb it with a Gaussian random walk
perturb_lambda <- function(lambda, Sigma_kernel) {

  Sigma_kernel <- make_positive_definite(Sigma_kernel, min_eig = 1e-8)

  v <- lambda_to_vec(lambda)
  v_new <- as.numeric(MASS::mvrnorm(1, mu = v, Sigma = Sigma_kernel))
  names(v_new) <- names(v)

  vec_to_lambda(v_new)
}

# Gaussian proposal density from the old particle to the new one
kernel_density <- function(lambda_new, lambda_old, Sigma_kernel) {
  x <- lambda_to_vec(lambda_new)
  m <- lambda_to_vec(lambda_old)
  exp(mvnorm_logpdf(x, m, Sigma_kernel))
}

# Diagnostics
effective_sample_size <- function(w) {
  1 / sum(w^2)
}

# ABC_SMC procedure
abc_smc <- function(S_obs, N, J, K, W, T_pop, alpha_quantile, pilot_size) {

  populations <- vector("list", T_pop)
  eps_vec <- numeric(T_pop)

  # Population 1: Initial Rejection Sampling
  cat("Population 1\n")

  # Pilot simulations to determine epsilon_1
  d_store <- numeric(pilot_size)

  for (m in seq_len(pilot_size)) {

    th <- sample_lambda()
    out_sim <- simulate_X_from_lambda(th, N, J, K)
    d_now <- distance_abc(summaries(out_sim$X), S_obs)

    d_store[m] <- d_now
  }

  eps_vec[1] <- as.numeric(quantile(d_store, probs = alpha_quantile, na.rm = TRUE))

  cat("epsilon_1 =", eps_vec[1], "\n")

  # Generate population 1
  particles <- vector("list", W)
  distances <- numeric(W)

  n_accept <- 0
  total_prop_1 <- 0

  while (n_accept < W) {

    th <- sample_lambda()
    out_sim <- simulate_X_from_lambda(th, N, J, K)
    d_now <- distance_abc(summaries(out_sim$X), S_obs)

    total_prop_1 <- total_prop_1 + 1

    if (d_now <= eps_vec[1]) {
      n_accept <- n_accept + 1
      particles[[n_accept]] <- th
      distances[n_accept] <- d_now
    }
  }

  weights <- rep(1 / W, W)

  populations[[1]] <- list(
    particles = particles,
    weights = weights,
    distances = distances,
    distances_for_epsilon = distances,
    epsilon = eps_vec[1],
    proposals = total_prop_1,
    ESS = effective_sample_size(weights)
  )

  cat(
    "accepted =", W,
    " proposals =", total_prop_1,
    " ESS =", effective_sample_size(weights),
    "\n"
  )

  # Populations 2 to T_pop
  if (T_pop >= 2) {

    for (t in 2:T_pop) {

      cat("Population", t, "\n")

      prev_particles <- populations[[t - 1]]$particles
      prev_weights <- populations[[t - 1]]$weights
      prev_distances <- populations[[t - 1]]$distances_for_epsilon

      eps_vec[t] <- as.numeric(quantile(prev_distances, probs = alpha_quantile, na.rm = TRUE))

      cat(paste0("epsilon_", t, " = "), eps_vec[t], "\n")

      # Construct the Gaussian perturbation kernel
      prev_mat <- do.call(rbind, lapply(prev_particles, lambda_to_vec))
      Sigma_kernel <- 2 * weighted_covariance(prev_mat, prev_weights)
      Sigma_kernel <- make_positive_definite(Sigma_kernel, min_eig = 1e-8)

      particles_t <- vector("list", W)
      distances_t <- numeric(W)
      weights_t <- numeric(W)

      n_accept <- 0
      total_prop_t <- 0

      while (n_accept < W) {

        # Select and perturb a previous particle
        idx <- sample(seq_len(W), size = 1, prob = prev_weights)
        th_star <- perturb_lambda(prev_particles[[idx]], Sigma_kernel)
        total_prop_t <- total_prop_t + 1

        if (!in_prior_support(th_star)) {
          next
        }

        # Simulate and calculate ABC distance
        out_sim <- simulate_X_from_lambda(th_star, N, J, K)
        d_now <- distance_abc(summaries(out_sim$X), S_obs)

        if (d_now <= eps_vec[t]) {

          num <- prior_density(th_star)
          denom <- sum(prev_weights * sapply(prev_particles, function(old) {
            kernel_density(th_star, old, Sigma_kernel)
          }))

          if (
            !is.finite(num) || num <= 0 ||
            !is.finite(denom) || denom <= 0
          ) {
            next
          }

          n_accept <- n_accept + 1
          particles_t[[n_accept]] <- th_star
          distances_t[n_accept] <- d_now
          weights_t[n_accept] <- num / denom
        }
      }

      # Normalize importance weights
      weights_t <- weights_t / sum(weights_t)
      ess_t <- effective_sample_size(weights_t)

      # Resample particles and corresponding distances
      accepted_distances_t <- distances_t
      if (ess_t < W / 2) {

        idx_resample <- sample(seq_len(W), size = W, replace = TRUE, prob = weights_t)
        particles_t <- particles_t[idx_resample]
        distances_t <- distances_t[idx_resample]
        weights_t <- rep(1 / W, W)

        ess_t <- effective_sample_size(weights_t)
      }

      populations[[t]] <- list(
        particles = particles_t,
        weights = weights_t,
        distances = distances_t,
        distances_for_epsilon = accepted_distances_t,
        epsilon = eps_vec[t],
        proposals = total_prop_t,
        Sigma_kernel = Sigma_kernel,
        ESS = ess_t
      )

      cat(
        "accepted =", W,
        " proposals =", total_prop_t,
        " ESS =", ess_t,
        "\n"
      )
    }
  }

  list(
    populations = populations,
    eps_vec = eps_vec
  )
}
```

```{r}
# Number of cells / genes / TFs
N_use <- nrow(X_obs)
J_use <- ncol(X_obs)

# Target accepted samples per population
W <- 200

# Number of ABC-SMC populations
T_pop <- 5

# Quantile used to update tolerances
alpha_quantile <- 0.25

# run ABC-SMC
smc_fit <- abc_smc(S_obs = S_obs, N = N_use, J = J_use, K = K_use, W = W, 
                   T_pop = T_pop, alpha_quantile = alpha_quantile, pilot_size = 2000)

# True prior sample for prior-vs-posterior plots
prior_size <- 2000

prior_draws <- lapply(1:prior_size, function(m) {
  lambda <- sample_lambda()
  c(theta = lambda$theta, gamma = lambda$gamma, rho = lambda$rho, alpha = lambda$alpha, 
    sigma2 = lambda$sigma2, alpha_phi = lambda$alpha_phi, beta_phi = lambda$beta_phi, 
    alpha_psi = lambda$alpha_psi, beta_psi = lambda$beta_psi)
})

prior_df <- as.data.frame(do.call(rbind, prior_draws))

# Final ABC-SMC population as posterior
last_pop <- smc_fit$populations[[T_pop]]
theta_acc <- last_pop$particles
w_acc <- last_pop$weights
d_acc <- last_pop$distances

if (any(abs(w_acc - 1 / length(w_acc)) > 1e-12)) {
  idx_post <- sample(seq_along(theta_acc), size = length(theta_acc),
                     replace = TRUE, prob = w_acc)

  theta_acc <- theta_acc[idx_post]
  d_acc <- d_acc[idx_post]
}

w_acc <- rep(1 / length(theta_acc), length(theta_acc))

post_df <- do.call(
  rbind,
  lapply(theta_acc, function(lambda) {
    c(
      theta = lambda$theta,
      gamma = lambda$gamma,
      rho = lambda$rho,
      alpha = lambda$alpha,
      sigma2 = lambda$sigma2,
      alpha_phi = lambda$alpha_phi,
      beta_phi = lambda$beta_phi,
      alpha_psi = lambda$alpha_psi,
      beta_psi = lambda$beta_psi
    )
  })
)
post_df <- as.data.frame(post_df)
post_df$w <- w_acc
post_df$d <- d_acc

cat("Prior size =", nrow(prior_df), "\n")
cat("Posterior size =", nrow(post_df), "\n")
```

```{r}
# Prior vs Posterior plots
op <- par(no.readonly = TRUE)

# Page 1: theta, log(gamma), log(rho)
par(mfrow = c(3, 2), mar = c(4, 4, 2.5, 1))

# 1) theta
x_theta_prior <- prior_df$theta
x_theta_post <- post_df$theta
xlim_theta <- range(c(x_theta_prior, x_theta_post), finite = TRUE)
br_theta <- seq(xlim_theta[1], xlim_theta[2], length.out = 61)

hist(x_theta_prior, breaks = br_theta, probability = TRUE,
     main = "Prior: gene_on probability (theta)", xlab = "theta", xlim = xlim_theta)
hist(x_theta_post,  breaks = br_theta, probability = TRUE,
     main = "Posterior: theta", xlab = "theta", xlim = xlim_theta)

# 2) log(gamma)
x_lg_prior <- log(prior_df$gamma)
x_lg_post <- log(post_df$gamma)
xlim_lgamma <- range(c(x_lg_prior, x_lg_post), finite = TRUE)
br_lgamma <- seq(xlim_lgamma[1], xlim_lgamma[2], length.out = 61)

hist(x_lg_prior, breaks = br_lgamma, probability = TRUE,
     main = "Prior: biological variability log(gamma)", xlab = "log(gamma)", xlim = xlim_lgamma)
hist(x_lg_post, breaks = br_lgamma, probability = TRUE,
     main = "Posterior: log(gamma)", xlab = "log(gamma)", xlim = xlim_lgamma)

# 3) log(rho)
x_lr_prior <- log(prior_df$rho)
x_lr_post <- log(post_df$rho)
xlim_lrho <- range(c(x_lr_prior, x_lr_post), finite = TRUE)
br_lrho <- seq(xlim_lrho[1], xlim_lrho[2], length.out = 61)

hist(x_lr_prior, breaks = br_lrho, probability = TRUE,
     main = "Prior: technical noise log(rho)", xlab = "log(rho)", xlim = xlim_lrho)
hist(x_lr_post,  breaks = br_lrho, probability = TRUE,
     main = "Posterior: log(rho)", xlab = "log(rho)", xlim = xlim_lrho)

# Page 2: log(alpha), log(sigma2), log(alpha_phi)
par(mfrow = c(3, 2), mar = c(4, 4, 2.5, 1))

# 4) log(alpha)
x_la_prior <- log(prior_df$alpha)
x_la_post <- log(post_df$alpha)
xlim_lalpha <- range(c(x_la_prior, x_la_post), finite = TRUE)
br_lalpha <- seq(xlim_lalpha[1], xlim_lalpha[2], length.out = 61)

hist(x_la_prior, breaks = br_lalpha, probability = TRUE,
     main = "Prior: TF activity variability log(alpha)", xlab = "log(alpha)", xlim = xlim_lalpha)
hist(x_la_post, breaks = br_lalpha, probability = TRUE,
     main = "Posterior: log(alpha)", xlab = "log(alpha)", xlim = xlim_lalpha)

# 5) log(sigma2)
x_ls_prior <- log(prior_df$sigma2)
x_ls_post <- log(post_df$sigma2)
xlim_lsigma2 <- range(c(x_ls_prior, x_ls_post), finite = TRUE)
br_lsigma2 <- seq(xlim_lsigma2[1], xlim_lsigma2[2], length.out = 61)

hist(x_ls_prior, breaks = br_lsigma2, probability = TRUE,
     main = "Prior: Regulatory effect variance log(sigma2)",
     xlab = "log(sigma2)", xlim = xlim_lsigma2)
hist(x_ls_post, breaks = br_lsigma2, probability = TRUE,
     main = "Posterior: log(sigma2)", xlab = "log(sigma2)", xlim = xlim_lsigma2)

# 6) log(alpha_phi)
x_lalpha_phi_prior <- log(prior_df$alpha_phi)
x_lalpha_phi_post <- log(post_df$alpha_phi)
xlim_lalpha_phi <- range(c(x_lalpha_phi_prior, x_lalpha_phi_post), finite = TRUE)
br_lalpha_phi <- seq(xlim_lalpha_phi[1], xlim_lalpha_phi[2], length.out = 61)

hist(x_lalpha_phi_prior, breaks = br_lalpha_phi, probability = TRUE,
     main = "Prior: Target Beta shape log(alpha_phi)", xlab = "log(alpha_phi)", 
     xlim = xlim_lalpha_phi)
hist(x_lalpha_phi_post, breaks = br_lalpha_phi, probability = TRUE,
     main = "Posterior: log(alpha_phi)", xlab = "log(alpha_phi)", xlim = xlim_lalpha_phi)

# Page 3: log(beta_phi), log(alpha_psi), log(beta_psi)
par(mfrow = c(3, 2), mar = c(4, 4, 2.5, 1))

# 7) log(beta_phi)
x_lbeta_phi_prior <- log(prior_df$beta_phi)
x_lbeta_phi_post <- log(post_df$beta_phi)
xlim_lbeta_phi <- range(c(x_lbeta_phi_prior, x_lbeta_phi_post), finite = TRUE)
br_lbeta_phi <- seq(xlim_lbeta_phi[1], xlim_lbeta_phi[2], length.out = 61)

hist(x_lbeta_phi_prior, breaks = br_lbeta_phi, probability = TRUE,
     main = "Prior: Target Beta shape log(beta_phi)", xlab = "log(beta_phi)", xlim = xlim_lbeta_phi)
hist(x_lbeta_phi_post, breaks = br_lbeta_phi, probability = TRUE,
     main = "Posterior: log(beta_phi)", xlab = "log(beta_phi)", xlim = xlim_lbeta_phi)

# 8) log(alpha_psi)
x_lalpha_psi_prior <- log(prior_df$alpha_psi)
x_lalpha_psi_post <- log(post_df$alpha_psi)
xlim_lalpha_psi <- range(c(x_lalpha_psi_prior, x_lalpha_psi_post), finite = TRUE)
br_lalpha_psi <- seq(xlim_lalpha_psi[1], xlim_lalpha_psi[2], length.out = 61)

hist(x_lalpha_psi_prior, breaks = br_lalpha_psi, probability = TRUE,
     main = "Prior: Regulator Beta shape log(alpha_psi)", xlab = "log(alpha_psi)", xlim = xlim_lalpha_psi)
hist(x_lalpha_psi_post, breaks = br_lalpha_psi, probability = TRUE,
     main = "Posterior: log(alpha_psi)", xlab = "log(alpha_psi)", xlim = xlim_lalpha_psi)

# 9) log(beta_psi)
x_lbeta_psi_prior <- log(prior_df$beta_psi)
x_lbeta_psi_post <- log(post_df$beta_psi)
xlim_lbeta_psi <- range(c(x_lbeta_psi_prior, x_lbeta_psi_post), finite = TRUE)
br_lbeta_psi <- seq(xlim_lbeta_psi[1], xlim_lbeta_psi[2], length.out = 61)

hist(x_lbeta_psi_prior, breaks = br_lbeta_psi, probability = TRUE,
     main = "Prior: Regulator Beta shape log(beta_psi)",
     xlab = "log(beta_psi)", xlim = xlim_lbeta_psi)
hist(x_lbeta_psi_post, breaks = br_lbeta_psi, probability = TRUE,
     main = "Posterior: log(beta_psi)", xlab = "log(beta_psi)", xlim = xlim_lbeta_psi)

par(op)
```

```{r}
# Posterior predictive (latent layers)
# Simulate from all accepted posterior thetas from the final ABC-SMC population
R_post <- length(theta_acc)
out_pp_list <- vector("list", R_post)

for (r in 1:R_post) {
  th_r <- theta_acc[[r]]
  out_pp_list[[r]] <- simulate_X_from_lambda(th_r, N_use, J_use, K_use)
}

X_pp_list <- lapply(out_pp_list, function(o) o$X)

# Latent variables
a_vec <- unlist(lapply(out_pp_list, function(o) as.double(c(o$a))), use.names = FALSE)
delta_vec <- unlist(lapply(out_pp_list, function(o) as.integer(c(o$delta))), use.names = FALSE)
beta_vec <- unlist(lapply(out_pp_list, function(o) as.double(c(o$beta))), use.names = FALSE)
m_vec <- unlist(lapply(out_pp_list, function(o) as.double(c(o$m))), use.names = FALSE)
pi_vec <- unlist(lapply(out_pp_list, function(o) as.double(o$pi_j)), use.names = FALSE)
phi_j_vec <- unlist(lapply(out_pp_list, function(o) as.double(o$phi_j)), use.names = FALSE)
psi_k_vec <- unlist(lapply(out_pp_list, function(o) as.double(o$psi_k)), use.names = FALSE)
v_vec <- unlist(lapply(out_pp_list, function(o) as.integer(c(o$v))), use.names = FALSE)
z_vec <- unlist(lapply(out_pp_list, function(o) as.double(c(o$z))), use.names = FALSE)
t_vec <- unlist(lapply(out_pp_list, function(o) as.double(c(o$t))), use.names = FALSE)

# 1) a_ik
a_vec <- a_vec[is.finite(a_vec)]
xr_a <- range(a_vec, na.rm = TRUE)
br_a <- seq(xr_a[1], xr_a[2], length.out = 101)

ggplot(data.frame(v = a_vec), aes(x = v)) +
  geom_histogram(aes(y = after_stat(count / sum(count))),
                 breaks = br_a, alpha = 0.4) +
  coord_cartesian(xlim = xr_a) +
  labs(title = "a_ik (TF activities)",
       x = "a_ik", y = "Probability") +
  theme_minimal()

# 2) delta_kj
delta_vec <- delta_vec[is.finite(delta_vec)]
tab_delta <- table(factor(delta_vec, levels = c(0, 1)))
df_delta <- data.frame(val = c("0", "1"), prob = as.numeric(tab_delta) / sum(tab_delta))

ggplot(df_delta, aes(x = val, y = prob)) +
  geom_col(alpha = 0.6) +
  labs(title = "delta_kj (adjacency)",
       x = "delta_kj", y = "Probability") +
  theme_minimal()

# 3) beta_kj
beta_vec <- beta_vec[is.finite(beta_vec)]
beta_nz <- beta_vec[beta_vec != 0]
xr_b <- range(beta_vec, na.rm = TRUE)
br_b <- seq(xr_b[1], xr_b[2], length.out = 101)

ggplot(data.frame(v = beta_vec), aes(x = v)) +
  geom_histogram(aes(y = after_stat(count / sum(count))),
                 breaks = br_b, alpha = 0.4) +
  coord_cartesian(xlim = xr_b) +
  labs(title = "beta_kj (regulatory effect)",
       x = "beta_kj", y = "Probability") +
  theme_minimal()

# beta_kj (nonzero)
if (length(beta_nz) > 0) {
  xr_bnz <- range(beta_nz, na.rm = TRUE)
  br_bnz <- seq(xr_bnz[1], xr_bnz[2], length.out = 101)

  ggplot(data.frame(v = beta_nz), aes(x = v)) +
    geom_histogram(aes(y = after_stat(count / sum(count))),
                   breaks = br_bnz, alpha = 0.4) +
    coord_cartesian(xlim = xr_bnz) +
    labs(title = "beta_kj (nonzero) (regulatory effect)",
         x = "beta_kj | != 0", y = "Probability") +
    theme_minimal()
}

# 4) m_ij
m_vec <- m_vec[is.finite(m_vec)]
xr_m <- range(m_vec, na.rm = TRUE)
br_m <- seq(xr_m[1], xr_m[2], length.out = 101)

ggplot(data.frame(v = m_vec), aes(x = v)) +
  geom_histogram(aes(y = after_stat(count / sum(count))),
                 breaks = br_m, alpha = 0.4) +
  coord_cartesian(xlim = xr_m) +
  labs(title = "m_ij (regulatory mean)",
       x = "m_ij", y = "Probability") +
  theme_minimal()

# 5) pi_j
pi_vec <- pi_vec[is.finite(pi_vec)]
xr_p <- range(pi_vec, na.rm = TRUE)
br_p <- seq(xr_p[1], xr_p[2], length.out = 61)

ggplot(data.frame(v = pi_vec), aes(x = v)) +
  geom_histogram(aes(y = after_stat(count / sum(count))),
                 breaks = br_p, alpha = 0.4) +
  coord_cartesian(xlim = xr_p) +
  labs(title = "pi_j (gene on-probability)",
       x = "pi_j", y = "Probability") +
  theme_minimal()

# 6) phi_j
phi_j_vec <- phi_j_vec[is.finite(phi_j_vec)]

ggplot(data.frame(v = phi_j_vec), aes(x = v)) +
  geom_histogram(
    aes(y = after_stat(count / sum(count))),
    breaks = seq(0, 1, length.out = 61),
    alpha = 0.4
  ) +
  coord_cartesian(xlim = c(0, 1)) +
  labs(
    title = "phi_j (gene-specific edge propensity)",
    x = "phi_j",
    y = "Probability"
  ) +
  theme_minimal()

# 7) psi_k
psi_k_vec <- psi_k_vec[is.finite(psi_k_vec)]

ggplot(data.frame(v = psi_k_vec), aes(x = v)) +
  geom_histogram(
    aes(y = after_stat(count / sum(count))),
    breaks = seq(0, 1, length.out = 61),
    alpha = 0.4
  ) +
  coord_cartesian(xlim = c(0, 1)) +
  labs(
    title = "psi_k (regulator-specific edge propensity)",
    x = "psi_k",
    y = "Probability"
  ) +
  theme_minimal()

# 8) v_ij bar plot
v_vec <- v_vec[is.finite(v_vec)]
tab_v <- table(factor(v_vec, levels = c(0, 1)))
df_v <- data.frame(val = c("0", "1"), prob = as.numeric(tab_v) / sum(tab_v))

ggplot(df_v, aes(x = val, y = prob)) +
  geom_col(alpha = 0.6) +
  labs(title = "v_ij (on/off)",
       x = "v_ij", y = "Probability") +
  theme_minimal()

# 9) z_ij
z_vec <- z_vec[is.finite(z_vec)]
xr_z <- range(z_vec, na.rm = TRUE)
br_z <- seq(xr_z[1], xr_z[2], length.out = 101)

ggplot(data.frame(v = z_vec), aes(x = v)) +
  geom_histogram(aes(y = after_stat(count / sum(count))),
                 breaks = br_z, alpha = 0.4) +
  coord_cartesian(xlim = xr_z) +
  labs(title = "z_ij (latent expression)",
       x = "z_ij", y = "Probability") +
  theme_minimal()

# 10) t_ij
t_vec <- t_vec[is.finite(t_vec)]
xr_t <- range(t_vec, na.rm = TRUE)
br_t <- seq(xr_t[1], xr_t[2], length.out = 101)

ggplot(data.frame(v = t_vec), aes(x = v)) +
  geom_histogram(aes(y = after_stat(count / sum(count))),
                 breaks = br_t, alpha = 0.4) +
  coord_cartesian(xlim = xr_t) +
  labs(title = "t_ij (technical multiplicative noise)",
       x = "t_ij", y = "Probability") +
  theme_minimal()
```

```{r}
# Posterior predictive distributions
# Entrywise vectors
x_obs_vec <- as.numeric(X_obs)
x_sim_vec <- unlist(lapply(X_pp_list, as.numeric), use.names = FALSE)

x_obs_nonzero <- x_obs_vec[x_obs_vec > 0]
x_sim_nonzero <- x_sim_vec[x_sim_vec > 0]

# Cell-level summaries
C_obs <- Matrix::rowSums(X_obs) / ncol(X_obs)
D_obs <- Matrix::rowSums(X_obs > 0) / ncol(X_obs)

C_sim <- unlist(lapply(X_pp_list, function(X) Matrix::rowSums(X) / ncol(X)), use.names = FALSE)
D_sim <- unlist(lapply(X_pp_list, function(X) Matrix::rowSums(X > 0) / ncol(X)), use.names = FALSE)

# Gene-level summaries
G_obs <- Matrix::colSums(X_obs) / nrow(X_obs)
M_obs <- Matrix::colSums(X_obs > 0) / nrow(X_obs)

G_sim <- unlist(lapply(X_pp_list, function(X) Matrix::colSums(X) / nrow(X)), use.names = FALSE)
M_sim <- unlist(lapply(X_pp_list, function(X) Matrix::colSums(X > 0) / nrow(X)), use.names = FALSE)

# Plot function
make_page_2x2_facet <- function(obs, sim, title_base, xlab_raw, xlab_log, bins = 100) {

  # log10(1+x) axis transform
  log10p1_trans <- scales::trans_new(
    name = "log10p1",
    transform = function(x) log10(x + 1),
    inverse = function(x) 10^x - 1,
    domain = c(-1 + 1e-12, Inf)
  )

  obs <- as.numeric(obs)
  sim <- as.numeric(sim)
  x_min <- min(c(obs, sim), na.rm = TRUE)
  x_max <- max(c(obs, sim), na.rm = TRUE)

  # histogram breaks
  br_raw <- seq(x_min, x_max, length.out = bins + 1)
  br_log_raw <- (10^seq(log10(1 + x_min), log10(1 + x_max), length.out = bins + 1)) - 1

  # axis ticks
  breaks_raw <- sort(unique(c(x_min, pretty(c(x_min, x_max), n = 5))))
  breaks_raw <- breaks_raw[breaks_raw >= x_min & breaks_raw <= x_max]

  log_min <- log10(1 + x_min); log_max <- log10(1 + x_max)
  breaks_log_space <- pretty(c(log_min, log_max), n = 5)
  breaks_log_raw <- (10^breaks_log_space) - 1
  breaks_log_raw <- sort(unique(c(x_min, breaks_log_raw)))
  breaks_log_raw <- breaks_log_raw[breaks_log_raw >= x_min & breaks_log_raw <= x_max]

  # probability
  hist_rect_df <- function(v, breaks, source) {
    h <- hist(v, breaks = breaks, plot = FALSE, right = FALSE)
    p <- if (sum(h$counts) > 0) h$counts / sum(h$counts) else rep(0, length(h$counts))
    data.frame(
      xmin = h$breaks[-length(h$breaks)],
      xmax = h$breaks[-1],
      prob = p,
      source = factor(source, levels = c("Observed", "Simulated"))
    )
  }

  df_raw <- rbind(
    hist_rect_df(obs, br_raw, "Observed"),
    hist_rect_df(sim, br_raw, "Simulated")
  )

  df_log <- rbind(
    hist_rect_df(obs, br_log_raw, "Observed"),
    hist_rect_df(sim, br_log_raw, "Simulated")
  )

  ymax_raw <- max(df_raw$prob, na.rm = TRUE)
  ymax_log <- max(df_log$prob, na.rm = TRUE)

  p_raw <- ggplot2::ggplot(df_raw) +
    ggplot2::geom_rect(ggplot2::aes(xmin = xmin, xmax = xmax, ymin = 0, ymax = prob), 
                       alpha = 0.4) +
    ggplot2::facet_grid(rows = ggplot2::vars(source), axes = "all", axis.labels = "all") +
    ggplot2::scale_x_continuous(
      limits = c(x_min, x_max),
      breaks = breaks_raw,
      labels = scales::label_number(big.mark = ","),
      expand = c(0, 0)
    ) +
    ggplot2::scale_y_continuous(
      breaks = function(lim) pretty(lim, n = 5),
      labels = scales::label_number(accuracy = 0.01),
      expand = c(0, 0)
    ) +
    ggplot2::coord_cartesian(ylim = c(0, ymax_raw)) +
    ggplot2::labs(title = paste0("RAW ", title_base), x = xlab_raw, y = "Probability") +
    ggplot2::theme_minimal()

  p_log <- ggplot2::ggplot(df_log) +
    ggplot2::geom_rect(ggplot2::aes(xmin = xmin, xmax = xmax, ymin = 0, ymax = prob), alpha = 0.4) +
    ggplot2::facet_grid(rows = ggplot2::vars(source), axes = "all", axis.labels = "all") +
    ggplot2::scale_x_continuous(
      limits = c(x_min, x_max),
      trans  = log10p1_trans,
      breaks = breaks_log_raw,
      labels = scales::label_number(big.mark = ","),
      expand = c(0, 0)
    ) +
    ggplot2::scale_y_continuous(
      breaks = function(lim) pretty(lim, n = 5),
      labels = scales::label_number(accuracy = 0.01),
      expand = c(0, 0)
    ) +
    ggplot2::coord_cartesian(ylim = c(0, ymax_log)) +
    ggplot2::labs(title = paste0("LOG ", title_base), x = xlab_log, y = "Probability") +
    ggplot2::theme_minimal()

  gridExtra::grid.arrange(p_raw, p_log, ncol = 2)
}

# 1) Entrywise all entries
make_page_2x2_facet(
  obs = x_obs_vec,
  sim = x_sim_vec,
  title_base = "Entrywise: counts (all)",
  xlab_raw = "X_ij",
  xlab_log = "X_ij",
  bins = 100
)

# 2) Entrywise positive entries only
make_page_2x2_facet(
  obs = x_obs_nonzero,
  sim = x_sim_nonzero,
  title_base = "Entrywise: counts (positive)",
  xlab_raw = "X_ij | X_ij > 0",
  xlab_log = "X_ij | X_ij > 0",
  bins = 100
)

# 3) Cell-level mean UMI per gene
make_page_2x2_facet(
  obs = as.numeric(C_obs),
  sim = as.numeric(C_sim),
  title_base = "Cell-level: mean UMI per gene",
  xlab_raw = "(1/J) sum_j X_ij",
  xlab_log = "(1/J) sum_j X_ij",
  bins = 100
)

# 4) Cell-level fraction detected genes
make_page_2x2_facet(
  obs = as.numeric(D_obs),
  sim = as.numeric(D_sim),
  title_base = "Cell-level: fraction detected genes",
  xlab_raw = "(1/J) sum_j 1{X_ij > 0}",
  xlab_log = "(1/J) sum_j 1{X_ij > 0}",
  bins = 100
)

# 5) Gene-level mean UMI per cell
make_page_2x2_facet(
  obs = as.numeric(G_obs),
  sim = as.numeric(G_sim),
  title_base = "Gene-level: mean UMI per cell",
  xlab_raw = "(1/N) sum_i X_ij",
  xlab_log = "(1/N) sum_i X_ij",
  bins = 100
)

# 6) Gene-level fraction expressing cells
make_page_2x2_facet(
  obs = as.numeric(M_obs),
  sim = as.numeric(M_sim),
  title_base = "Gene-level: fraction expressing cells",
  xlab_raw = "(1/N) sum_i 1{X_ij > 0}",
  xlab_log = "(1/N) sum_i 1{X_ij > 0}",
  bins = 100
)
```

```{r}
range(x_obs_vec, finite = TRUE)
range(x_sim_vec, finite = TRUE)
```
